Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-09-06 fstats_pca_lewontin.tex
revisions

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
Most genetic variation in humans is shared between all of us, but around 15\% of genetic variation in humans can be explained by population structure \citep{dobzhansky1972, barbujani1997, rosenberg2002a}. Genome-scale data has allowed us to leverage the information contained in these 15\% to study our genetic diversity and history in great detail \citep{cavalli-sforza1994,reich2018a}. For some data sets it is possible to predict an individuals origin at a resolution of a few hundred kilometers \citep{novembre2008, leslie2015}, and direct-to-consumer-genetics companies are using this variation to analyze the genetic data of millions of customers.

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies. Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical grounds, but also based on genetic data. Based on 15 blood group loci, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002a, reich2018a}, and we now know that human races cannot be motivated genetically, but rather are a polyphyletic grouping that is maintained by historical and social conventions.

Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}. In isolation, populations are expected to slowly diverge, particularly if they are separated by barriers to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}. On the other hand, major population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}. Local migration between neighboring populations will flatten differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010},will lead to locally increased diversity.

The interplay of these demographic processes leads to complex genetic structure that is difficult to model using standard population genetic models: Human populations are typically non-discrete, not at equilibrium and not randomly mating. One analysis paradigm is to use different tools based on different sets of assumptions; each emphasizing a particular aspects of the data. Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable. For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests, but with the drawback that strong assumptions about models need to be assumed.

Particularly in the analysis of human ancient DNA, a set of techniques based on $F$-statistics \textit{sensu} Patterson have risen in popularity \citep{patterson2012, peter2016}. This framework is based on the null-assumption that the relationship between  populations is tree-like, and then uses tests involving three or four populations to probe for violations of treeness \citep{patterson2012}. In a further step, $F$-statistics of multiple populations can be combined to estimate parameters in more complex models \citep{patterson2012, harney2021}. However, the connections between PCA, $F$-statistics and explicit demographic models are often unclear, which makes quantitative comparisons, detecting model violations and joint interpretation of these approaches difficult. Since both $F$-statistics and PCA are functions of expected pairwise coalescent times \citep{mcvean2009, peter2016}, this is one avenue to link these approaches. Here, I instead use the geometric interpretation of $F$-statistics developed by  \cite{oteo-garcia2021} to directly visualize $F$-statistics on a PCA plot.

\section{Theory}
In this section, I will give a very brief formal introduction to $F$-statistics and PCA. A more detailed technical treatise of PCA is given by e.g. \cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

\subsection{Introduction to PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, all results are valid if we treat each individual as its own population. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

@@

\subsection{Introduction to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

@@

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_on_pca.pdf}
	\caption{\textbf{Geometric representation of $F$-statistics on 2D-PCA-plot.} A: $F_2$ represents the squared Euclidean distance between two points in PC-space. B: Admixture-$F_3(X_x; X_1, X_2)$ is negative if $X_x$ lies in the circle specified by the diameter $X_2-X_1$. C: $F_3(X_x; X_1, X_2)$ is negative given $X_1, X_x$ if $X_2$ is in the gray space.  D: Outgroup-$F_3$ reflects the projection of $X_2 - X_O$ on $X_1 - X_O$. E: $F_4$ is the projection of $X_3 - X_4$ on $X_1-X_2$. F: Same projection, but assuming data is generated by a tree.}
	\label{fig:geom}
\end{figure}

>>>>

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\normsq}[1]{\left\lVert#1\right\rVert^2}
\newcommand{\MX}{\mathbf{X}} %uncentered data
\newcommand{\MC}{\mathbf{C}} %centering
\newcommand{\MY}{\mathbf{Y}} %centered data
\newcommand{\MF}{\mathbf{F}_2} %F2-distance matrix
\newcommand{\MFT}{\mathbf{F}_3} %F3-distance matrix
\newcommand{\MP}{\mathbf{P}} % PCs
\newcommand{\ML}{\mathbf{L}} % loadings
\newcommand{\MK}{\mathbf{K}} % Kernel
\newcommand{\MSINGULAR}{\mathbf{\Sigma}} % Singular values matrix
\newcommand{\MEIGEN}{\mathbf{\Lambda}} % Eigenvalue matrix
\newcommand{\MEAN}{\boldsymbol{\mu}} % Kernel
\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics.
Here, I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation.
\end{abstract}
\section{Introduction}
As in other species, the genetic diversity of human populations  is shaped by their environment and history over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994, schraiber2015}. In turn, studying the diversity patterns in humans can be used to reconstruct the demographic and evolutionary history of our species. As such, population genetic theory is essential to understand how this works.

\paragraph{processes}
Over time, genetic drift will slowly change allele frequencies and as a result, isolated populations are expected to slowly diverge. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance. However, isolation-by-distance patterns are not uniform, but shaped by geography, particularly barriers   to migration such as mountain ranges, oceans or deserts \citep{bradburd2013, peter2020a, rosenberg2005}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, peter2020a, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016} or secondary contact and admixture between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity.

Modern descriptions of human genetic diversity focus on the evolutionary processes that caused it, and which gives rise to  both discrete and continuous components \citep{rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a}.
The interplay of these demographic processes leads to the complex genetic structure that we observe in present-day people, and which we aim to reconstruct from genetic data, which is challenging since we cannot expect to devise a single model that captures all complexity.

\paragraph{Data-driven approaches}
One commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions; each emphasizing a particular aspects of the data. Particularly for large data sets with several thousand samples, exploratory, data-driven methods such as population trees, Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are a first step aimed at providing a simple, low-dimensional summary of the data while making minimal assumptions about the underlying population structure.

\paragraph{Model-based appraoches}
While PCA and related tools give a good overview of the data, they are not designed to answer specific research questions and cannot be used for model testing or parameter estimates. For this purpose, methods based on explicit demographic models  are often used \citep{gutenkunst2009, kamm2015, excoffier2013}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modelling assumptions such as that populations are discrete, randomly mating, and at equilibrium, with the hope that despite these assumptions being violated, the resulting model fits provide a sufficiently accurate description to answer the research questions.

\paragraph{F-stats}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models including all data can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset. However, this has the drawback that a large proportion of the data may be unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. I safe the formal definition for later; but the easiest way to motivate them assumes that populations are related as a tree, where the edges measure how much genetic drift has occurred. In this case, $F_2$ measures the genetic distance between populations, $F_3$ corresponds to an external, and $F_4$ to an internal branch of a tree ) \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, these $F$-statistics are estimated from some data, and then used as tests of treeness. The canonical alternative model is an admixture graph or phylogenetic network \citep{patterson2012}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). Typically, analyses proceed by testing specific hypotheses abot gene flow using $F$-statistics \citep{reich2009, green2010, lazaridis2014}. Increasingly commonly, $F$-statistics are also incorporated into tools that jointly fit more complicated models \citep[e.g. ][]{patterson2012,harney2021}. However, admixture graphs are not the only alternative model that explain violations of treeness. Expected $F$-statistics can be calculated for a wide range of demographic models \citep{peter2016}, or even in abstract Euclidean spaces \citep{oteo-garcia2021}, so that model fit remains a major concern.

\paragraph{PCA}
With minimal assumptions, PCA is one of the most widely used data summary and data exploration techniques. One way to think about PCA is that it aims to create a low-dimensional summary of the data that retains a maximum of information about the underlying structure in the data (Again, a formal description will be given in the next section). It has been widely used and independently discovered in many disciplines. In population genetics, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize global genetic diversity in the pre-genomic era \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale population genetic data set.  This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. Theoretically, the PCA-decomposition has been studied in terms of trees \citep{cavalli-sforza1975}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}.

\paragraph{F-stats and PCA}
However, the connections between PCA and $F$-statistics remains poorly understood, and the goal of this paper is to investigate this connection. To do so, I use the geometric interpretation of $F$-statistics developed by \cite{oteo-garcia2021} who defined them in arbitrary Euclidean spaces. We make use of the empirical finding that while the allele-frequency space is frequently high-dimensional, for many empirical data sets, a low number of dimensions are sufficient to characterize population structure, and hence $F$-statistics. This allows for approximate interpretations of $F$-statistics on PCA plots, and I use two sample data sets to evaluate these approximations in practice.

\section{Theory}
In this section, I will introduce the mathematics and notations for $F$-statistics and PCA. A comprehensive  treatise on PCA is given by e.g. \cite{jolliffe2013} and \cite{pachter2014}, and a useful guide to interpretation is \cite{cavalli-sforza1994}. Readers unfamiliar with $F$-statistics may find \cite{patterson2012}, \cite{peter2016} or \cite{oteo-garcia2021} helpful.

\subsection{Formal Definition of PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, all results are valid if we treat each individual as its own population. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

@@

\subsection{Formal Definition to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

@@

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
	\caption{\textbf{Representation of $F$-statistics on tree and 2D-PCA-plot.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot. A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space. B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthoginal projection of $X_1 - X_3$ onto $X_1$-$X_4$. C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$. D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
	\label{fig:geom}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_admixture_pca.pdf}
	\caption{\textbf{Admixture representation on 2D-PCA-plot.} The schematics show four populations and their representation using an admixture graph (A) or a 2D-PCA plot. A: Admixture graph, with population $X_y$ originating as an admixture of $X_2$ and $X_3$, with $X_2$ contributing proportion $\alpha$. Subsequent drift (red branch) will change allele frequency to sampled admixture population $X_x$. B: PCA representation of the scenario in A. $X_y$ originates on the segment connecting $X_2$ and $X_3$, and subsequent drift may move it in a random direction. C: $F_3(X_x; X_2, X_3)$ and negative region (light gray circle). $F_4(X_1, X_x; X_3, X_4)$ will no longer be zero (compare to Figure \ref{fig:geom}D). }
	\label{fig:admix}
\end{figure}
